TEMPERATURES AND METALLICITIES FOR 10,000+ M DWARFS IN THE \apogee\ SURVEY

### 2018-08-11 paper/draft.tex
some math text fixes

\newcommand{\apogee}{\textsl{APOGEE}}
\newcommand{\thecannon}{\textsl{The Cannon}}
\newcommand{\cannon}{\textsl{Cannon}}
\newcommand{\aspcap}{\textsl{ASPCAP}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\zmass}{\textsl{2MASS}}
\newcommand{\sdssiv}{\textsl{SDSS-IV}}
\newcommand{\teff}{T_{\rm eff}}

@@

\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; near-infrared) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H]) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ we train a two-parameter model precise to 68K and 0.07 dex respectively. We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that \thecannon\ can infer spectral types to a precision of 0.6 types. Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in the survey and compare our results to \apogee\ pipeline measurements, and find our results to be significantly more precise and accurate than \aspcap, with scatters of $\sim80-100$K comapred to color-temperature relations from several literature sources.

@@

The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).

@@

The physical parameter sample consists of 87 sources with reference labels distributed over $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15. The M15 catalog in total contains 183 sources and the extension sample another 500 primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog. All targets have low-res optical spectra from the SNIFS specgrograph \citep{Lantz:2004} and infrared spectra taken with the SpeX Spectograph \citep{Rayner:2003}, which have been combined to estaimte largely empirical bolometric fluxes. Effective temperatures have been estimated by comparing the SNIFS spectra to BT-SETTL atmospheric models \citep{Allard:2011}. A subsample of 29 sources with measured angular diameters from long-baseline optical interferometry \citep{Boyajian:2012} are used to calibrate the model comparison, including masking of spectral regions poorly reproduced by the model spectra \citep{Mann:2013c}. Based on difference between assigned \teff\ values and those from angular diameters, absolute uncertainty on \teff\ is esitmated to be 60K in T$_{\rm eff}$, although the relative uncertainty is likely a factor of $\simeq$2 better.
%AWM: I realized I've never actaully sat down and rigorously measured the relative Teff uncertainty. The results from the Cannon actually suggest ~30 is not bad. I just did some quick tests, and I'm not prepared to put an absolute number on it. A \simeq 30 here should be safeish.

>>>>

\newcommand{\apogee}{\textsl{APOGEE}}
\newcommand{\thecannon}{\textsl{The Cannon}}
\newcommand{\cannon}{\textsl{Cannon}}
\newcommand{\aspcap}{\textsl{ASPCAP}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\zmass}{\textsl{2MASS}}
\newcommand{\sdssiv}{\textsl{SDSS-IV}}
\newcommand{\teff}{$T_{\rm eff}$}

@@

\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; near-infrared) spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H]) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim$3550K \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ we train a two-parameter model precise to 68K and 0.07 dex respectively. We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that \thecannon\ can infer spectral types to a precision of 0.6 types. Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in the survey and compare our results to \apogee\ pipeline measurements, and find our results to be significantly more precise and accurate than \aspcap, with scatters of $\sim80-100$K comapred to color-temperature relations from several literature sources.

@@

The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (\teff, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).

@@

The physical parameter sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15. The M15 catalog in total contains 183 sources and the extension sample another 500 primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog. All targets have low-res optical spectra from the SNIFS specgrograph \citep{Lantz:2004} and infrared spectra taken with the SpeX Spectograph \citep{Rayner:2003}, which have been combined to estaimte largely empirical bolometric fluxes. Effective temperatures have been estimated by comparing the SNIFS spectra to BT-SETTL atmospheric models \citep{Allard:2011}. A subsample of 29 sources with measured angular diameters from long-baseline optical interferometry \citep{Boyajian:2012} are used to calibrate the model comparison, including masking of spectral regions poorly reproduced by the model spectra \citep{Mann:2013c}. Based on difference between assigned \teff\ values and those from angular diameters, absolute uncertainty on \teff\ is esitmated to be 60K in \teff, although the relative uncertainty is likely a factor of $\simeq$2 better.
%AWM: I realized I've never actaully sat down and rigorously measured the relative Teff uncertainty. The results from the Cannon actually suggest ~30 is not bad. I just did some quick tests, and I'm not prepared to put an absolute number on it. A \simeq 30 here should be safeish.
